A matrix-free approach for finite-strain hyperelastic problems using geometric multigrid

### 2019-01-11 doc/large_strain.tex
Update doc/large_strain.tex
Co-Authored-By: davydden <davydden@gmail.com>
update on what we measure as memory, att todo's

\usepackage{bm}
\usepackage{xcolor}
\usepackage{listings}
\lstset{ %
   backgroundcolor=\color{white},   % choose the background color; you must add \usepackage{color} or \usepackage{xcolor}
   basicstyle=\small,        % the size of the fonts that are used for the code
   breakatwhitespace=false,         % sets if automatic breaks should only happen at whitespace
   breaklines=false,                 % sets automatic line breaking
   captionpos=b,                    % sets the caption-position to bottom
   %commentstyle=\color{mygreen},    % comment style
   deletekeywords={...},            % if you want to delete keywords from the given language
   escapeinside={\%*}{*)},          % if you want to add LaTeX within your code
   extendedchars=false,              % lets you use non-ASCII characters; for 8-bits encodings only, does not work with UTF-8
   frame=single,	                   % adds a frame around the code
   keepspaces=true,                 % keeps spaces in text, useful for keeping indentation of code (possibly needs columns=flexible)
   keywordstyle=\color{blue},       % keyword style
   language=C++,                 % the language of the code
   otherkeywords={*,...},           % if you want to add more keywords to the set
   numbers=left,                    % where to put the line-numbers; possible values are (none, left, right)
   numbersep=5pt,                   % how far the line-numbers are from the code
   numberstyle=\tiny\color{gray}, % the style that is used for the line-numbers
   rulecolor=\color{black},         % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. comments (green here))
   showspaces=false,                % show spaces everywhere adding particular underscores; it overrides 'showstringspaces'
   showstringspaces=false,          % underline spaces within strings only
   showtabs=false,                  % show tabs within strings adding particular underscores
   stepnumber=2,                    % the step between two line-numbers. If it is 1, each line will be numbered
%   stringstyle=\color{mymauve},     % string literal style
   tabsize=2,	                   % sets default tabsize to 2 spaces
   title=\lstname                   % show the filename of files included with \lstinputlisting; also try caption instead of title
}
\usepackage[linesnumbered,lined,commentsnumbered,ruled]{algorithm2e}

@@

{\color{red}
To summarize, in this section we propose a method by which to construct level tangent operators for finite-strain elasticity to be employed within the GMG preconditioner.
The other parts of the GMG preconditioner such as level smoothers, matrix-free interlevel transfer operators and the coarse level iterative solver are provided by the \texttt{deal.II} library.
The efficiency of the preconditioner, from the perspective of linear algebra, with the so defined level operators will be assessed by numerical examples in the next section.
We will, however, refrain from studying the machine performance aspects of the multigrid preconditioner as a whole as the focus of this paper is the matrix-free implementation of tangent operators of finite-strain elasticity both at the fine mesh level as well as the coarser multigrid levels.
Finally, we note that at the time of writing, $p-$ version of the GMG is not available within the \texttt{deal.II} library.
}

@@

Figures \ref{fig:benchmark_miehe_Emmy} and \ref{fig:benchmark_miehe_IWR} show the results of the matrix-vector multiplication for the considered finite-strain hyperelastic benchmark problem as performed on the two clusters.
{\color{red}
We are interested in the wall-clock time (average over 10 runs) per DoF as a first metric.}
As expected, the matrix-vector multiplication becomes very expensive for higher order discretization for sparse matrix-based approaches,
{\color{red}as the matrix becomes more and more dense.
The performance results for finite-strain elasticity operators considered in this study are qualitatively similar to those reported for Laplace operator in \cite{kronbichler12}.
Note that we consider only non-Cartesian meshes together with additional mappings (namely, manifold descriptions of the boundaries and interfaces to capture the geometry as well as
the linearization-related mapping required for evaluation of gradients with respect to the deformed configuration).
Both factors increase the memory consumption and make element operations more expensive.
Nonetheless }
the matrix-free implementation is faster than the matrix-based counterpart already for cubic Lagrangian basis in 2D and quadratic in 3D.
Figures \ref{fig:benchmark_miehe_Emmy_vmult2}, \ref{fig:benchmark_miehe_Emmy_vmult3}, \ref{fig:benchmark_miehe_IWR_vmult2} and \ref{fig:benchmark_miehe_IWR_vmult3} clearly show the influence of the different caching strategies,
{\color{red}
described in Algorithms \ref{alg:mf_scalar}, \ref{alg:mf_tensor2} and \ref{alg:mf_tensor4}.}
Two conclusions can be drawn from these results:
The scalar caching implementation is the most time consuming of the three, as at each step we additionally evaluate the gradient of the displacement field in the referential configuration that is required to evaluate the Kirchhoff stress $\gz \tau$.
There is little difference {\color{red}in terms of the wall-clock time per DoF} between the approach where the material part of the fourth-order spatial tangent stiffness tensor is cached (``tensor4'') and the one where we only cache the second-order Kirchhoff tensor and utilize the chosen hyperelastic constitutive equation to efficiently implement the action of the material part of the fourth-order spatial tangent stiffness tensor on the second-order symmetric tensor.
This indicates that the time savings we get from the latter approach are insignificant within the complete matrix-free operator implementation.
Consequently, we can apply the matrix-free approach to any finite-strain material model by caching the material part of the fourth-order spatial tangent stiffness tensor in addition to the Kirchhoff stress and expect this to be faster than the matrix-based implementation.
It is therefore feasible to define highly performant generic operators that are independent of any applied constitutive laws,
{\color{red}so long as the resulting tangent operator is expressed via the \eqref{eq:algebraic_tangent}.}

As outlined in the introduction, it is not only the performance of the matrix-free vector multiplication, but also the memory requirement to store a sparse matrix which is the driving force behind matrix-free methods.
Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3} demonstrate that even with caching one fourth-order symmetric tensor and one second-order tensor at each quadrature point, the matrix-free approach takes much less memory than its matrix-based counterpart.

>>>>

\usepackage{todonotes}
\usepackage{bm}
\usepackage{xcolor}
\usepackage{listings}
\lstset{ %
   backgroundcolor=\color{white},   % choose the background color; you must add \usepackage{color} or \usepackage{xcolor}
   basicstyle=\small,        % the size of the fonts that are used for the code
   breakatwhitespace=false,         % sets if automatic breaks should only happen at whitespace
   breaklines=false,                 % sets automatic line breaking
   captionpos=b,                    % sets the caption-position to bottom
   %commentstyle=\color{mygreen},    % comment style
   deletekeywords={...},            % if you want to delete keywords from the given language
   escapeinside={\%*}{*)},          % if you want to add LaTeX within your code
   extendedchars=false,              % lets you use non-ASCII characters; for 8-bits encodings only, does not work with UTF-8
   frame=single,	                   % adds a frame around the code
   keepspaces=true,                 % keeps spaces in text, useful for keeping indentation of code (possibly needs columns=flexible)
   keywordstyle=\color{blue},       % keyword style
   language=C++,                 % the language of the code
   otherkeywords={*,...},           % if you want to add more keywords to the set
   numbers=left,                    % where to put the line-numbers; possible values are (none, left, right)
   numbersep=5pt,                   % how far the line-numbers are from the code
   numberstyle=\tiny\color{gray}, % the style that is used for the line-numbers
   rulecolor=\color{black},         % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. comments (green here))
   showspaces=false,                % show spaces everywhere adding particular underscores; it overrides 'showstringspaces'
   showstringspaces=false,          % underline spaces within strings only
   showtabs=false,                  % show tabs within strings adding particular underscores
   stepnumber=2,                    % the step between two line-numbers. If it is 1, each line will be numbered
%   stringstyle=\color{mymauve},     % string literal style
   tabsize=2,	                   % sets default tabsize to 2 spaces
   title=\lstname                   % show the filename of files included with \lstinputlisting; also try caption instead of title
}
\usepackage[linesnumbered,lined,commentsnumbered,ruled]{algorithm2e}

@@

{\color{red}
To summarize, in this section we propose a method by which to construct level tangent operators for finite-strain elasticity to be employed within the GMG preconditioner.
The other parts of the GMG preconditioner such as level smoothers, matrix-free interlevel transfer operators and the coarse level iterative solver are provided by the \texttt{deal.II} library.
The efficiency of the preconditioner, from the perspective of linear algebra, with the so defined level operators will be assessed by numerical examples in the next section.
We will, however, refrain from studying the machine performance aspects of the multigrid preconditioner as a whole as the focus of this paper is the matrix-free implementation of tangent operators of finite-strain elasticity both at the fine mesh level as well as the coarser multigrid levels.
Finally, we note that at the time of writing, $p-$ version of the GMG is not available within the \texttt{deal.II} library and therefore not considered here.
}

@@

Figures \ref{fig:benchmark_miehe_Emmy} and \ref{fig:benchmark_miehe_IWR} show the results of the matrix-vector multiplication for the considered finite-strain hyperelastic benchmark problem as performed on the two clusters.
{\color{red}
We are interested in the wall-clock time (average over 10 runs per each linear solver step) per DoF as a first metric.}
As expected, the matrix-vector multiplication becomes very expensive for higher order discretization for sparse matrix-based approaches,
{\color{red}as the matrix becomes more and more dense.
The performance results for finite-strain elasticity operators considered in this study are qualitatively similar to those reported for Laplace operator in \cite{kronbichler12}.
Note that we consider only non-Cartesian meshes together with additional mappings (namely, manifold descriptions of the boundaries and interfaces to capture the geometry as well as
the linearization-related mapping required for evaluation of gradients with respect to the deformed configuration).
Both factors increase the memory consumption and make element operations more expensive.
Nonetheless }
the matrix-free implementation is faster than the matrix-based counterpart already for cubic Lagrangian basis in 2D and quadratic in 3D.
Figures \ref{fig:benchmark_miehe_Emmy_vmult2}, \ref{fig:benchmark_miehe_Emmy_vmult3}, \ref{fig:benchmark_miehe_IWR_vmult2} and \ref{fig:benchmark_miehe_IWR_vmult3} clearly show the influence of the different caching strategies,
{\color{red}
described in Algorithms \ref{alg:mf_scalar}, \ref{alg:mf_tensor2} and \ref{alg:mf_tensor4}.}
Two conclusions can be drawn from these results:
The scalar caching implementation is the most time consuming of the three, as at each step we additionally evaluate the gradient of the displacement field in the referential configuration that is required to evaluate the Kirchhoff stress $\gz \tau$.
There is little difference {\color{red}in terms of the wall-clock time per DoF} between the approach where the material part of the fourth-order spatial tangent stiffness tensor is cached (``tensor4'') and the one where we only cache the second-order Kirchhoff tensor and utilize the chosen hyperelastic constitutive equation to efficiently implement the action of the material part of the fourth-order spatial tangent stiffness tensor on the second-order symmetric tensor.
This indicates that the time savings we get from the latter approach are insignificant within the complete matrix-free operator implementation.
Consequently, we can apply the matrix-free approach to any finite-strain material model by caching the material part of the fourth-order spatial tangent stiffness tensor in addition to the Kirchhoff stress and expect this to be faster than the matrix-based implementation.
It is therefore feasible to define highly performant generic operators that are independent of any applied constitutive laws,
{\color{red}so long as the resulting tangent operator is expressed via the \eqref{eq:algebraic_tangent}.}

As outlined in the introduction, it is not only the performance of the matrix-free vector multiplication, but also the memory requirement to store a sparse matrix which is the driving force behind matrix-free methods.
Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3} demonstrate that even with caching one fourth-order symmetric tensor and one second-order tensor at each quadrature point, the matrix-free approach takes much less memory than its matrix-based counterpart.
{\color{red}
In those graphs we report the memory needed to cache additional quantities required for the tangent operator here considered,
as well as the memory required by the \texttt{deal.II} \textit{MatrixFree} object, which caches the inverse of the Jacobian of the transformation
from unit to real cell \cite{kronbichler12}.
Note that for the Algorithm \ref{alg:mf_scalar} we need to evaluate gradients with respect to both deform and undeformed
configurations, and therefore two inverses will be cached via two different \textit{MatrixFree} objects.\todo{rerun RRZE and IWR}\todo{this is roughly main memory transfer, do we elaborate on this/use it?}}
